\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-01-17 paper/main.tex
inputting a round of written edits

\title{\Large Carryover effects in free recall reveal how past experiences influence
memories of future experiences}

@@

\begin{abstract} \footnotesize{We perceive, interpet, and remember ongoing experiences
through the lens of our prior experiences. Inferring that we are in one type of
situation versus another can lead us to interpret the same physical experience
differently. In turn, this can affect how we focus our attention, form
expectations about what will happen next, remember what is happening now, draw on
our prior related experiences, and so on. To study these phenomena, we asked
participants to perform simple word list learning tasks. Across different
experimental conditions, we held the set of to-be-learned words constant, but
we manipulated the orders in which the words were studied. We found that these
order manipulations affected not only how the participants recalled the ordered
lists, but also how they recalled later randomly ordered lists. Our work shows
how structure in our ongoing experiences can exert influence on how we remember
unrelated subsequent experiences.

@@

The contexts we encounter help us to construct \textit{situation
models}~\citep{RangRitc12, MannEtal15} or \textit{schemas}~\citep{MasiEtal22,
BaldEtal18} that describe how experiences are likely to unfold based on our
prior experiences with similar contextual cues. For example, when we enter a
sit-down restaurant, we might expect to be seated at a table, given a menu, and
served food. Priming someone to expect a particular situation or context can
also influence how they resolve potential ambiguities in their ongoing
experiences, including ambiguous movies and narratives~\citep{YeshEtal17}.

@@

Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in part
by substantial differences in how we study these processes. Situation models
and schemas are most often studied using ``naturalistic'' stimuli such as
narratives and movies~\citep{ZwaaEtal95,ZwaaRadv98, NastEtal20}. In contrast,
our understanding of how we organize our memories has been most widely studied
using more traditional paradigms like free recall of random word
lists~\citep{Kaha12, Kaha20}. In free recall, participants study lists of items and are
instructed to recall the items in any order they choose. The orders in which
words come to mind can provide insights into how participants have organized
their memories of the studied words. Because random word lists are unstructured
by design, it is not clear if or how non-trivial situation models might apply
to these stimuli. Nevertheless, there are \textit{some} commonalities between
memory for word lists and memory for real-world experiences.

Like remembering real-world experiences, remembering words on a studied list
requires distinguishing the current list from the rest of one's experience. To
model this fundamental memory capability, cognitive scientists have posited a
special context representation that is associated with each list. According to
early theories~\citep[e.g.][]{Este55a,AndeBowe72} context representations are
composed of many features which fluctuate from moment to moment, slowly
drifting through a multidimensional feature space. During recall, this
representation forms part of the retrieval cue, enabling us to distinguish list
items from non-list items. Understanding the role of context in memory
processes is particularly important in self-cued memory tasks, such as
\textit{free recall}, where the retrieval cue is ``context'' itself.
Conceptually, the same general processes might be said to describe how
real-world contexts evolve during natural experiences. However, this is still
an open area of study~\citep{Mann20, Mann21a}.

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, SedeEtal08, PolyEtal09,
ShanHowa12}. These phenomena include the well-known recency and primacy effects
(superior recall of items from the end and, to a lesser extent, from the
beginning of the study list), as well as semantic and temporal clustering
effects~\citep{KahaEtal08b}. The contiguity effect is an example of temporal
clustering, which is perhaps the dominant form of organization in free recall.
This effect can be seen in the tendency for people to successively recall items
that occupied neighboring positions in the study list~\citep{Kaha96}. There are
also striking effects of semantic clustering~\citep{RomnEtal93, Bous53,
BousEtal54, JenkRuss52, MannKaha12}, whereby the recall of a given item is more
likely to be followed by recall of a similar or related item than a dissimilar
or unrelated one. In general, people organize memories for words along a wide
variety of stimulus dimensions. As formalized by models like the \textit{Context
Maintenance and Retrieval Model}~\citep{PolyEtal09}, the stimulus features
associated with each word (e.g.\ the word's meaning, font size, font color,
location on the screen, size of the object the word represents, etc.) are
incorporated into the participant's mental context
representation~\citep{SmitVela01, MannEtal15, Mann20, MannEtal11, MannEtal12}.
During a memory test, any of these features may serve as a memory cue, which in
turn leads the participant to recall in succession words that share stimulus
features.

@@

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). Across the different conditions in the experiment, we
varied the lists' presentation orders in different ways across lists. The
studied items (words) were designed to vary along three general dimensions:
semantic (word \textit{category}, and physical \textit{size} of the referent),
lexicographic (word \textit{length} and \textit{first letter}), and visual
(font \textit{color} and the onscreen \textit{location} of each word). In our
main manipulation conditions, we asked participants to study and recall eight
lists whose items were sorted by a target feature (e.g., word category). Next,
we asked them to study and recall an additional eight lists whose items had the
same features, but that were sorted in a random temporal order. We were
interested in how these order manipulations affected participants' recall
behaviors on early (sorted) lists, as well as how order manipulations on early
lists affected recall behaviors on later (unsorted) lists. We used a series of
control conditions as a baseline; in these control conditions all of the lists
were sorted randomly, but we manipulated the presence or absence of the visual
features. Finally, in an \textit{adaptive} experimental condition we used
participants' recall behaviors on early lists to manipulate, in real-time, the
presentation orders of subsequent lists. In this adaptive condition we varied
the agreement between how participants preferred to organize their memories of
the studied items versus the orders in which the items were presented.

@@

We enrolled a total of 491 Dartmouth undergraduate students across 11
experimental conditions. The conditions included two primary controls (feature
rich, reduced), two secondary controls (reduced (early), reduced (late)), six
order manipulation conditions (category, size, length, first letter, color, and
location), and a final adaptive condition. Each of these conditions are
described in the \textit{Experimental design} subsection below.

@@

Participants were assigned to experimental conditions based loosely on their
date of participation. (This aspect of our procedure helped us to more easily
synchronize the experiment databases across multiple testing computers.) Of the
490 participants who opted to fill out the demographics survey, reported ages
ranged from 17 to 31 years (mean: 19.1 years; standard deviation: 1.356 years).
A total of 318 participants reported their gender as female, 170 as male, and
two participants declined to report their gender. A total of 442 participants
reported their ethnicity as ``not Hispanic or Latino,'' 39 as ``Hispanic or
Latino,'' and nine declined to report their ethnicity. Participants reported
their races as White (345 participants), Asian (120 participants), Black or
African American (31 participants), American Indian or Alaska Native (11
participants), Native Hawaiian or Other Pacific Islander (four participants),
Mixed race (three participants), Middle Eastern (one participant), and Arab
(one participant). A total of five participants declined to report their race.
We note that several participants reported more than one of racial category.
Participants reported their highest degrees achieved as ``Some college'' (359
participants), ``High school graduate'' (117 participants), ``College
graduate'' (seven participants), ``Some high school'' (five participants),
``Doctorate'' (one participant), and ``Master's degree'' (one participant). A
total of 482 participants reported no reading impairments, and eight reported
having mild reading impairments. A total of 489 participants reported having
normal color vision and one participant reported that they were red-green color
blind. A total of 482 participants reported taking no prescription medications
and having no recent injuries; four participants reported having ADHD, one reported
having dyslexia, one reported having allergies, one reported a recently torn
ACL/MCL, and one reported a concussion from several months prior. The
participants reported consuming 0 -- 3 cups of coffee prior to the testing
session (mean: 0.32 cups; standard deviation: 0.58 cups). Participants reported
their current level of alertness, and we converted their responses to numerical
scores as follows: ``very sluggish'' (-2), ``a little sluggish'' (-1),
``neutral'' (0), ``a little alert'' (1), and ``very alert'' (2). Across all
participants, the full range of alertness levels were reported (range: -2 -- 2;
mean: 0.35; standard deviation: 0.89).

We dropped from our dataset the one participant who reported having abnormal
color vision, as well as 39 participants whose data were corrupted due to
technical failures while running the experiment or during the daily database
merges. In total, this left usable data from 452 participants, broken down by
experimental condition as follows: feature rich (67 participants), reduced (61
participants), reduced (late) (41 participants), reduced (early), (42
participants), category (30 participants), size (30 participants), length (30
participants), first letter (30 participants), color (31 participants),
location (30 participants), and adaptive (60 participants). The participant who
declined to fill out their demographic survey participated in the location
condition, and we verified verbally that they had normal color vision and no
significant reading impairments.

@@

\caption{\textbf{Feature-rich free recall.} After studying lists comprised of
words that vary along several feature dimensions, participants verbally recall
words in any order (microphone icon). Each experimental condition manipulates
word features and/or presentation orders within and/or across lists. The rows
display representative (illustrated) examples of the first lists participants
might encounter in each condition. The rectangles during the ``Presentation
phase'' show illustrated screen captures during a series of word presentations.
Each word appeared onscreen for 2 seconds, followed by 2 seconds of blank
screen. The red microphone icons during the ``Recall'' phase denote the one
minute verbal recall interval. The labels on the right (and corresponding
groupings on the left) denote experimental condition labels.}

@@

We assigned the categorized words into a total of 16 lists with several
constraints. First, we required that each list contained words from exactly 4
unique categories, each with exactly 4 exemplars from each category. Second, we
required that (across all words on the list) at least one instance of both
object sizes were represented. On average, each category was represented in
4.27 lists (standard deviation: 1.16 lists). Aside from these two constraints,
we assigned each word to a unique list. After random assignment, each list
contained words with an average of 11.13 unique starting letters (standard
deviation: 1.15 letters) and an average word length of 6.17 letters (standard
deviation: 0.34 letters).

@@

To select a random font color for each word, we drew three integers uniformly
and at random from the interval $\left[0, 255\right]$, corresponding to the red (r),
green (g), and blue (b) color channels for that word. To assign random
presentation locations to each word, we selected two floating point numbers
uniformly at random (one for the word's horizontal $x$ coordinate and the other
for its vertical $y$ coordinate). The bounds of these coordinates were selected
to cover the entire visible area of the display without cutting off any part of
the words. The words were shown on 27~in (diagonal) Retina 5K iMac displays
(resolution: 5120 $\times$ 2880 pixels).

@@

We used four ``control'' conditions to evaluate and explore participants'
baseline behaviors. We also used performance on these control conditions to
help interpret performance in other ``manipulation'' conditions. Two control
conditions served as ``anchor points.'' In the first anchor point condition,
which we call the \textit{feature rich} condition, we randomly shuffled the
presentation order (independently for each participant) of the words on each
list. In the second anchor point condition, which we call the \textit{reduced}
condition, we randomized word presentations as in the feature rich condition.
However, rather than assigning each word a random color and location, we
instead displayed all of the words in black and at the center of the screen.

In the \textit{reduced (early)} condition, we followed the ``reduced''
procedure (presenting each word in black at the center of the screen) for early
lists, and followed the ``feature rich'' procedure (presenting each word in a
random color and location) for late lists. Finally, in the \textit{reduced
(late)} condition, we followed the feature rich procedure for early lists and
the reduced procedure for late lists.

@@

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of features. Semantic features (category and size)
are \textit{categorical}. For these features, we defined a binary distance
function: two words were considered to ``match'' (i.e., have a distance of 0)
if their labels are the same (i.e., both from the same semantic category or
both of the same size). If two words' labels were different for a given
feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and words starting
with `J' and `A' respectively would have a first letter distance of 9. Because
words' lengths and letters' positions in the alphabet are always integers,
these discrete distances always take on integer values. Finally, the visual
features (color and location) are \textit{continuous} and
\textit{multivariate}, in that each ``feature'' takes on multiple (positive)
real values. We defined the ``color'' and ``location'' distances between two
words as the Euclidean distances between their $(r, g, b)$ color or $(x, y)$
location vectors, respectively. Therefore the color and location distance
measures always take on positive real values (upper-bounded at 441.67 for
color, or 27~in for location, reflecting the distances between the
corresponding maximally different vectors).

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique lists for each participant.
First, we choose a word uniformly at random from the set of candidates. Next,
we compute the distances between the chosen word's feature(s) and the
corresponding feature(s) of all yet-to-be-presented words. Third, we convert
these distances (between the previously presented word's feature values, $a$,
and the candidate word's feature values, $b$) to similarity scores:
\begin{equation} \mathrm{similarity}(a, b) = \exp\{-\tau \cdot
\mathrm{distance}(a, b)\}, \label{eqn:similarity} \end{equation} where $\tau =
1$ in our implementation. We note that increasing the value of $\tau$ would
amplify the influence of similarity on order, and decreasing the value of
$\tau$ would diminish the influence of similarity on order. Also note that this
approach requires $\tau > 0$. Finally, we computed a set of normalized
similarity values by dividing the similarities by their sum: \begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, \end{equation} where in the
denominator, $i$ takes on each of the $n$ feature values of the to-be-presented
words. The resulting set of normalized similarity scores sums to one.

@@

\caption{\textbf{Generating stochastic feature-sorted lists.} For a given
feature dimension (e.g., color), we compute the similarity
(Eqn.~\ref{eqn:similarity}) between the feature value(s) of the previous item,
$x$, and all yet-to-be-presented items ($a$ -- $g$). Next, we normalize these
similarity scores so that they sum to one. We lay, in sequence, a set of
``sticks,'' one for each candidate item, whose lengths are equal to these
normalized similarity scores. Note that the combined lengths of these sticks is
one. To select the next to-be-presented item, we draw a random number, $r$,
from the uniform distribution bounded between 0 and 1 (inclusive). The identity
of the next item is given by the stick adjacent to an indicator that moves
distance $r$ (starting from 0) along the sequence of sticks. In this case, the
next to-be-presented item is $e$. Note that each item's chances of selection is
proportional to its similarity to the previous item, along the given feature
dimension.} \label{fig:stick}

@@

As illustrated in Figure~\ref{fig:stick}, we use these normalized similarity
scores to construct a sequence of ``sticks'' that we lay end to end in a line.
Each of the $n$ sticks corresponds to a single to-be-presented word, and the
stick lengths are proportional to the relative similarities between each word's
feature value(s) and the feature value(s) of the just-presented word. We choose
the next to-be-presented word by moving an indicator along the set of sticks,
by a distance chosen uniformly at random on the interval $\left[0, 1\right]$.
We select the word associated with the stick lying next to the indicator to be
presented next. This process continues iteratively (re-computing the similarity
scores and stochastically choosing the next to-be-presented word using the
just-presented word) until all of the words have been presented. The result is
an ordered list that tends to change gradually along the selected feature
dimension.

@@

Temporal clustering describes a participant's tendency to organize their recall
sequences by the learned items' encoding positions. For instance, if a
participant recalled the lists' words in the exact order they were presented
(or in exact reverse order), this would yield a score of 1. If a participant
recalled the words in random order, this would yield an expected score of 0.5.
For each recall transition (and separately for each participant), we sorted all
not-yet-recalled words according to their absolute lag (that is, distance away
in the list). We then computed the percentile rank of the next word the
participant recalled. We took an average of these percentile ranks across all
of the participant’s recalls to obtain a single (uncorrected) temporal
clustering score for the participant.

@@

Suppose that two lists contain unequal numbers of items of each size. For
example, suppose that list $A$ contains all ``large'' items, whereas list $B$
contains an equal mix of ``large'' and ``small'' items. For a participant
recalling list $A$, any correctly recalled item will necessarily match the size
of the previous correctly recalled item. In other words, successively recalling
several list $A$ items of the same size is essentially meaningless, since
\textit{any} correctly recalled list $A$ word will be large. In contrast,
successively recalling several list $B$ items \textit{could} be meaningful,
since (early in the recall sequence) the yet-to-be-recalled items come from a
mix of sizes. However, once all of the small items on list $B$ have been
recalled, the best possible next matching recall will be a large item. And all
subsequent correct recalls must also be large items-- so for those later
recalls it becomes difficult to determine whether the participant is
successively recalling large items because they are organizing their
memories according to size, or (alternatively), whether they are simply
recalling the yet-to-be-recalled items in a random order. In general, the
precise order and blend of feature values expressed in a given list, the orders
and numbers of correct recalls a participant makes, the number of intervening
presentation positions between successive recalls, and so on, can all affect
the range of clustering scores that are possible to observe for a given list.
An uncorrected clustering score therefore conflates participants' actual
memory organization with other ``nuisance'' factors.

@@

Across six order manipulation conditions, we sorted early lists by each feature
dimension but randomly ordered the items on late lists (see \textit{Order
manipulation conditions}; features: category, size, length, first letter,
color, and location). Participants in the category-ordered condition showed an
increase in memory performance on early lists (accuracy, relative to early
feature rich lists; $t(95) = 3.034, p = 0.003$). Participants in the

@@

= 0.067$). Participants' performance on early lists in all of the other order
manipulation conditions was indistinguishable from performance on the early

>>>>

\title{\Large Feature and order manipulations in a free recall task affect memory for current and future lists}

@@

\begin{abstract} \footnotesize{We perceive, interpet, and remember ongoing
experiences through the lens of our prior experiences. Inferring that we are in
one type of situation versus another can lead us to interpret the same physical
experience differently. In turn, this can affect how we focus our attention,
form expectations about what will happen next, remember what is happening now,
draw on our prior related experiences, and so on. To study these phenomena, we
asked participants to perform simple word list learning tasks. Across different
experimental conditions, we held the set of to-be-learned words constant, but
we manipulated how irrelevant visual features changed across words and lists,
along with the orders in which the words were studied. We found that these
manipulations affected not only how the participants recalled the manipulated
lists, but also how they recalled later (random) lists. Our work shows how
structure in our ongoing experiences can exert influence over how we remember
our current experiences and unrelated subsequent experiences.

@@

The contexts we encounter help us to construct \textit{situation
models}~\citep{RangRitc12, MannEtal15, RadvCope06, ZwaaEtal95} or
\textit{schemas}~\citep{MasiEtal22, BaldEtal18} that describe how experiences
are likely to unfold based on our prior experiences with similar contextual
cues. For example, when we enter a sit-down restaurant, we might expect to be
seated at a table, given a menu, and served food. Priming someone to expect a
particular situation or context can also influence how they resolve potential
ambiguities in their ongoing experiences, including in ambiguous movies and
narratives~\citep{YeshEtal17}.

@@

Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in part
by substantial differences in how we study these processes. Situation models
and schemas are most often studied using ``naturalistic'' stimuli such as
narratives and movies~\citep{ZwaaEtal95,ZwaaRadv98, NastEtal20}. In contrast,
our understanding of how we organize our memories has been most widely informed
by more traditional paradigms like free recall of random word
lists~\citep{Kaha12, Kaha20}. In free recall, participants study lists of items
and are instructed to recall the items in any order they choose. The orders in
which words come to mind can provide insights into how participants have
organized their memories of the studied words. Because random word lists are
unstructured by design, it is not clear if or how non-trivial situation models
might apply to these stimuli. Nevertheless, there are \textit{some}
commonalities between memory for word lists and memory for real-world
experiences.

Like remembering real-world experiences, remembering words on a studied list
requires distinguishing the current list from the rest of one's experience. To
model this fundamental memory capability, cognitive scientists have posited a
special context representation that is associated with each list. According to
early theories~\citep[e.g.][]{Este55a,AndeBowe72} context representations are
composed of many features which fluctuate from moment to moment, slowly
drifting through a multidimensional feature space. During recall, this
representation forms part of the retrieval cue, enabling us to distinguish list
items from non-list items. Understanding the role of context in memory
processes is particularly important in self-cued memory tasks, such as free
recall, where the retrieval cue is ``context'' itself. Conceptually, the same
general processes might be said to describe how real-world contexts evolve
during natural experiences. However, this is still an open area of
study~\citep{Mann20, Mann21a}.

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, SedeEtal08, PolyEtal09,
ShanHowa12}. These phenomena include the well-known recency and primacy effects
(superior recall of items from the end and, to a lesser extent, from the
beginning of the study list), as well as semantic and temporal clustering
effects~\citep{KahaEtal08b}. The contiguity effect is an example of temporal
clustering, which is perhaps the dominant form of organization in free recall.
This effect can be seen in the tendency for people to successively recall items
that occupied neighboring positions in the studied list~\citep{Kaha96}. There
are also striking effects of semantic clustering~\citep{RomnEtal93, Bous53,
BousEtal54, JenkRuss52, MannKaha12}, whereby the recall of a given item is more
likely to be followed by recall of a similar or related item than a dissimilar
or unrelated one. In general, people organize memories for words along a wide
variety of stimulus dimensions. As formalized by models like the
\textit{Context Maintenance and Retrieval Model}~\citep{PolyEtal09}, the
stimulus features associated with each word (e.g.\ the word's meaning, font
size, font color, location on the screen, size of the object the word
represents, etc.) are incorporated into the participant's mental context
representation~\citep{SmitVela01, MannEtal15, Mann20, MannEtal11, MannEtal12}.
During a memory test, any of these features may serve as a memory cue, which in
turn leads the participant to recall in succession words that share stimulus
features.

@@

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). Across the different conditions in the experiment, we
varied the lists' appearances and presentation orders in different ways across
lists. The studied items (words) were designed to vary along three general
dimensions: semantic (word \textit{category}, and physical \textit{size} of the
referent), lexicographic (word \textit{length} and \textit{first letter}), and
visual (font \textit{color} and the onscreen \textit{location} of each word).
In two conditions, we manipulated whether the words' appearances were fixed or
variable within each list. In six manipulation conditions, we asked
participants to study and recall eight lists whose items were sorted by a
target feature (e.g., word category). Next, we asked them to study and recall
an additional eight lists whose items had the same features, but that were
sorted in a random temporal order. We were interested in how these 
manipulations affected participants' recall behaviors on early (manipulated) lists,
as well as how order manipulations on early lists affected recall behaviors on
later (random) lists. We used two control conditions as a baseline;
in these control conditions all of the lists were sorted randomly, but we
manipulated the presence or absence of the visual features. Finally, in an
\textit{adaptive} experimental condition we used participants' recall behaviors
on early lists to manipulate, in real-time, the presentation orders of
subsequent lists. In this adaptive condition we varied the agreement between
how participants preferred to organize their memories of the studied items
versus the orders in which the items were presented.

@@

We enrolled a total of 491 Dartmouth undergraduate students across 11
experimental conditions. The conditions included two controls (feature rich,
reduced), two visual manipulation conditions (reduced (early) and reduced
(late)), six order manipulation conditions (category, size, length, first
letter, color, and location), and a final adaptive condition. Each of these
conditions are described in the \textit{Experimental design} subsection below.

@@

Participants were assigned to experimental conditions based loosely on their
date of participation. (This aspect of our procedure helped us to more easily
synchronize the experiment databases across multiple testing computers.) Of the
490 participants who opted to fill out the demographics survey, reported ages
ranged from 17 to 31 years (mean: 19.1 years; standard deviation: 1.356 years).
A total of 318 participants reported their gender as female, 170 as male, and
two participants declined to report their gender. A total of 442 participants
reported their ethnicity as ``not Hispanic or Latino,'' 39 as ``Hispanic or
Latino,'' and nine declined to report their ethnicity. Participants reported
their races as White (345 participants), Asian (120 participants), Black or
African American (31 participants), American Indian or Alaska Native (11
participants), Native Hawaiian or Other Pacific Islander (four participants),
Mixed race (three participants), Middle Eastern (one participant), and Arab
(one participant). A total of five participants declined to report their race.
We note that several participants reported more than one of the above racial
categories. Participants reported their highest degrees achieved as ``Some
college'' (359 participants), ``High school graduate'' (117 participants),
``College graduate'' (seven participants), ``Some high school'' (five
participants), ``Doctorate'' (one participant), and ``Master's degree'' (one
participant). A total of 482 participants reported no reading impairments, and
eight reported having mild reading impairments. A total of 489 participants
reported having normal color vision and one participant reported that they were
red-green color blind. A total of 482 participants reported taking no
prescription medications and having no recent injuries; four participants
reported having ADHD, one reported having dyslexia, one reported having
allergies, one reported a recently torn ACL/MCL, and one reported a concussion
from several months prior. The participants reported consuming 0 -- 3 cups of
coffee prior to the testing session (mean: 0.32 cups; standard deviation: 0.58
cups). Participants reported their current level of alertness, and we converted
their responses to numerical scores as follows: ``very sluggish'' (-2), ``a
little sluggish'' (-1), ``neutral'' (0), ``a little alert'' (1), and ``very
alert'' (2). Across all participants, the full range of alertness levels were
reported (range: -2 -- 2; mean: 0.35; standard deviation: 0.89).

We dropped from our dataset the one participant who reported having abnormal
color vision, as well as 39 participants whose data were corrupted due to
technical failures while running the experiment or during the daily database
merges. In total, this left usable data from 452 participants, broken down by
experimental condition as follows: feature rich (67 participants), reduced (61
participants), reduced (early), (42 participants), reduced (late) (41
participants), category (30 participants), size (30 participants), length (30
participants), first letter (30 participants), color (31 participants),
location (30 participants), and adaptive (60 participants). The participant who
declined to fill out their demographic survey participated in the location
condition, and we verified verbally that they had normal color vision and no
significant reading impairments.

@@

\caption{\textbf{Feature-rich free recall.} After studying lists comprised of
words that vary along several feature dimensions, participants verbally recall
words in any order (microphone icon). Each experimental condition manipulates
word features and/or presentation orders within and/or across lists. The rows
display representative (illustrated) examples of items from the first lists
participants might encounter in each condition. The rectangles during the
``Presentation phase'' show illustrated screen captures during a series of word
presentations. Each word appeared onscreen for 2 seconds, followed by 2 seconds
of blank screen. The red microphone icons during the ``Recall'' phase denote
the one minute verbal recall interval. The labels on the right (and
corresponding groupings on the left) denote experimental condition labels.}

@@

We assigned the categorized words into a total of 16 lists with several
constraints. First, we required that each list contained words from exactly
four unique categories, each with exactly four exemplars from each category.
Second, we required that (across all words on the list) at least one instance
of both object sizes were represented. On average, each category was
represented in 4.27 lists (standard deviation: 1.16 lists). Aside from these
two constraints, we assigned each word to a unique list. After random
assignment, each list contained words with an average of 11.13 unique starting
letters (standard deviation: 1.15 letters) and an average word length of 6.17
letters (standard deviation: 0.34 letters).

@@

To select a random font color for each word, we drew three integers uniformly
and at random from the interval $\left[0, 255\right]$, corresponding to the red
(r), green (g), and blue (b) color channels for that word. To assign random
presentation locations to each word, we selected two floating point numbers
uniformly and at random (one for the word's horizontal $x$ coordinate and the
other for its vertical $y$ coordinate). The bounds of these coordinates were
selected to cover the entire visible area of the display without cutting off
any part of the words. The words were shown on 27~in (diagonal) Retina 5K iMac
displays (resolution: 5120 $\times$ 2880 pixels).

@@

We used two ``control'' conditions to evaluate and explore participants'
baseline behaviors. We also used performance on these control conditions to
help interpret performance in other ``manipulation'' conditions. In the first
control condition, which we call the \textit{feature rich} condition, we
randomly shuffled the presentation order (independently for each participant)
of the words on each list. In the second control condition, which we call the
\textit{reduced} condition, we randomized word presentations as in the feature
rich condition. However, rather than assigning each word a random color and
location, we instead displayed all of the words in black and at the center of
the screen.

We also designed two conditions where we varied the words' visual appearnaces
across lists. In the \textit{reduced (early)} condition, we followed the
``reduced'' procedure (presenting each word in black at the center of the
screen) for early lists, and followed the ``feature rich'' procedure
(presenting each word in a random color and location) for late lists. Finally,
in the \textit{reduced (late)} condition, we followed the feature rich
procedure for early lists and the reduced procedure for late lists.

@@

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of feature under consideration. Semantic features
(category and size) are \textit{categorical}. For these features, we defined a
binary distance function: two words were considered to ``match'' (i.e., have a
distance of 0) if their labels were the same (i.e., both from the same semantic
category or both of the same size). If two words' labels were different for a
given feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and words starting
with `J' and `A' respectively would have a first letter distance of 9. Because
words' lengths and letters' positions in the alphabet are always integers,
these discrete distances always take on integer values. Finally, the visual
features (color and location) are \textit{continuous} and
\textit{multivariate}, in that each ``feature'' takes on multiple (positive)
real values. We defined the ``color'' and ``location'' distances between two
words as the Euclidean distances between their $(r, g, b)$ color or $(x, y)$
location vectors, respectively. Therefore the color and location distance
measures always take on non-negative real values (upper-bounded at 441.67 for
color, or 27~in for location, reflecting the distances between the
corresponding maximally different vectors).

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique orderings for each
participant. First, we choose a word uniformly and at random from the set of
candidates. Next, we compute the distances between the chosen word's feature(s)
and the corresponding feature(s) of all yet-to-be-presented words. Third, we
convert these distances (between the previously presented word's feature
values, $a$, and the candidate word's feature values, $b$) to similarity
scores: \begin{equation} \mathrm{similarity}(a, b) = \exp\{-\tau \cdot
\mathrm{distance}(a, b)\}, \label{eqn:similarity} \end{equation} where $\tau =
1$ in our implementation. We note that increasing the value of $\tau$ would
amplify the influence of similarity on order, and decreasing the value of
$\tau$ would diminish the influence of similarity on order. Also note that this
approach requires $\tau > 0$. Finally, we computed a set of normalized
similarity values by dividing the similarities by their sum: \begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, \end{equation} where in the
denominator, $i$ takes on each of the $n$ feature values of the to-be-presented
words. The resulting set of normalized similarity scores sums to 1.

@@

\caption{\textbf{Generating stochastic feature-sorted lists.} For a given
feature dimension (e.g., color), we compute the similarity
(Eqn.~\ref{eqn:similarity}) between the feature value(s) of the previous item,
$x$, and all yet-to-be-presented items ($a$ -- $g$). Next, we normalize these
similarity scores so that they sum to 1. We lay, in sequence, a set of
``sticks,'' one for each candidate item, whose lengths are equal to these
normalized similarity scores. To select the next to-be-presented item, we draw
a random number, $r$, from the uniform distribution bounded between 0 and 1
(inclusive). The identity of the next item is given by the stick adjacent to an
indicator that moves distance $r$ (starting from 0) along the sequence of
sticks. In this case, the next to-be-presented item is $e$. Note that each
item's chances of selection is proportional to its similarity to the previous
item, along the given feature dimension.} \label{fig:stick}

@@

As illustrated in Figure~\ref{fig:stick}, we use these normalized similarity
scores to construct a sequence of ``sticks'' that we lay end to end in a line.
Each of the $n$ sticks corresponds to a single to-be-presented word, and the
stick lengths are proportional to the relative similarities between each word's
feature value(s) and the feature value(s) of the just-presented word. We choose
the next to-be-presented word by moving an indicator along the set of sticks,
by a distance chosen uniformly and at random on the interval $\left[0,
1\right]$. We select the word associated with the stick lying next to the
indicator to be presented next. This process continues iteratively
(re-computing the similarity scores and stochastically choosing the next
to-be-presented word using the just-presented word) until all of the words have
been presented. The result is an ordered list that tends to change gradually
along the selected feature dimension.

@@

Temporal clustering describes a participant's tendency to organize their recall
sequences by the learned items' encoding positions. For instance, if a
participant recalled the lists' words in the exact order they were presented
(or in exact reverse order), this would yield a score of 1. If a participant
recalled the words in a random order, this would yield an expected score of
0.5. For each recall transition (and separately for each participant), we
sorted all not-yet-recalled words according to their absolute lag (that is,
distance away in the list). We then computed the percentile rank of the next
word the participant recalled. We took an average of these percentile ranks
across all of the participant’s recalls to obtain a single (uncorrected)
temporal clustering score for the participant.

@@

Suppose that two lists contain unequal numbers of items of each size. For
example, suppose that list $A$ contains all ``large'' items, whereas list $B$
contains an equal mix of ``large'' and ``small'' items. For a participant
recalling list $A$, any correctly recalled item will necessarily match the size
of the previous correctly recalled item. In other words, successively recalling
several list $A$ items of the same size is essentially meaningless, since
\textit{any} correctly recalled list $A$ word will be large. In contrast,
successively recalling several list $B$ items of the same size \textit{could}
be meaningful, since (early in the recall sequence) the yet-to-be-recalled
items come from a mix of sizes. However, once all of the small items on list
$B$ have been recalled, the best possible next matching recall will be a large
item. And all subsequent correct recalls must also be large items-- so for
those later recalls it becomes difficult to determine whether the participant
is successively recalling large items because they are organizing their
memories according to size, or (alternatively), whether they are simply
recalling the yet-to-be-recalled items in a random order. In general, the
precise order and blend of feature values expressed in a given list, the orders
and numbers of correct recalls a participant makes, the number of intervening
presentation positions between successive recalls, and so on, can all affect
the range of clustering scores that are possible to observe for a given list.
An uncorrected clustering score therefore conflates participants' actual memory
organization with other ``nuisance'' factors.

@@

Across each of six order manipulation conditions, we sorted early lists by one
feature dimension but randomly ordered the items on late lists (see
\textit{Order manipulation conditions}; features: category, size, length, first
letter, color, and location). Participants in the category-ordered condition
showed an increase in memory performance on early lists (accuracy, relative to
early feature rich lists; $t(95) = 3.034, p = 0.003$). Participants in the

@@

= 0.067$). Participants' performances on early lists in all of the other order
manipulation conditions were indistinguishable from performance on the early
